---
title: "01: Quality Control"
output:
  github_document:
    html_preview: false
---

## R Libraries

The following R libraries were used to process the data and produce this report.

```{r libraries, results=FALSE, message=FALSE, warning=FALSE}
library(dplyr)
library(Seurat)
library(stringr)
library(ggplot2)
library(SummarizedExperiment)
library(openxlsx)
library(scDblFinder)
library(SoupX)
library(data.table)
library(scds)
library(magrittr)
```

---

```{r params, results=FALSE, message=FALSE, warning=FALSE}
sampleNames <- c(
  "EC1", "EC2", "EC2_FU", "EC3", "EC4",
  "NV1", "NV2", "NV3", "NV4", "NV5", "NV6",
  "NV7", "NV8", "NV9", "NV10", "NV11", "CTRL2",
  "CTRL3", "CTRL4", "CTRL5", "CTRL7"
)

data_dir    <- "../MEX"
projectName <- "CAR-T LC"

# QC thresholds
nFeature_min  <- 200
nFeature_max  <- 8000
percent_mt_max <- 20
nCount_min     <- 1000      # exceptions below
nCount_max     <- 100000
nCount_CTRL3   <- 3500      # higher minimum for CTRL3, based on per sample distribution. See line 292
nCount_NV5   <- 1650      # higher minimum for NV5, based on per sample distribution. See line 292
```

---

## Ambient RNA correction (SoupX)

```{r soupx_loop, message=FALSE, warning=FALSE}
set.seed(2024)

# Check if all files are getting picked from directory of interest
all_dirs <- list.dirs(path = data_dir, full.names = TRUE, recursive = FALSE)
message("Detected directories: ", paste(all_dirs, collapse = ", "))

seurat_list <- list()

pdf("pdfs/raw_SoupX_diagnostic_plots.pdf", width = 8, height = 6)

for (sample in sampleNames) {

  filtered_file <- file.path(data_dir, sample)
  raw_file      <- file.path(data_dir, paste0(sample, "_raw"))

  message("Processing sample: ", sample)
  message("raw: ", raw_file)
  message("filtered: ", filtered_file)

  tod <- Read10X(raw_file)
  toc <- Read10X(filtered_file)

  # Build SoupChannel and estimate soup profile
  sc <- SoupChannel(tod, toc, calcSoupProfile = FALSE)
  sc <- estimateSoup(sc)

  # clustering for SoupX 
  tmp <- CreateSeuratObject(counts = toc, min.cells = 0, min.features = 0)
  tmp <- NormalizeData(tmp)
  tmp <- FindVariableFeatures(tmp, nfeatures = 2000)
  tmp <- ScaleData(tmp)
  tmp <- RunPCA(tmp, npcs = 30)
  tmp <- RunUMAP(tmp, dims = 1:30)
  tmp <- FindNeighbors(tmp, dims = 1:30)
  tmp <- FindClusters(tmp)

  sc <- setClusters(sc, setNames(tmp$seurat_clusters, rownames(tmp@meta.data)))
  sc <- setDR(sc, tmp@reductions$umap@cell.embeddings)

  plot.new()
  title(main = paste("SoupX diagnostic —", sample), cex.main = 1.2)
  sc <- autoEstCont(sc)

  # Corrected
  corrected <- adjustCounts(sc, roundToInt = TRUE)

  # Build final Seurat object from corrected counts
  seurat_obj <- CreateSeuratObject(
    counts    = corrected,
    project   = projectName,
    min.cells = 3,
    min.features = 200
  )
  seurat_obj$orig.ident <- sample

  toc_matched <- toc[, colnames(corrected)]
  seurat_obj[["RNA"]]$original_counts <- toc_matched

  seurat_list[[sample]] <- seurat_obj
}

dev.off()

# Merge all samples into a single Seurat object
combined_data <- merge(seurat_list[[1]], y = seurat_list[-1], add.cell.ids = sampleNames)
combined_data$sample_names <- combined_data$orig.ident
```

```{r condition_labels, message=FALSE, warning=FALSE}
# CAR-T Enterocolitis: Patients who received CAR-T and developed colitis
# CAR-T Control: Patients who received CAR-T but no colitis
# NV: Healthy controls
# EC2: Follow-up sample

sample_to_condition <- setNames(
  ifelse(sampleNames == "EC2_FU", "EC2_FU",
    ifelse(sampleNames %in% c("CTRL2", "CTRL3", "CTRL4", "CTRL5", "CTRL7"),
           "CAR-T Control",
      ifelse(grepl("^EC", sampleNames), "CAR-T Enterocolitis",
        ifelse(grepl("^NV",   sampleNames), "NV", NA)
      )
    )
  ),
  sampleNames
)

combined_data$condition <- plyr::mapvalues(
  combined_data$sample_names,
  from = names(sample_to_condition),
  to   = sample_to_condition
)
unique(combined_data$condition)
```

```{r category,  message=FALSE, warning=FALSE}
sample_to_category <- setNames(
  case_when(
    
    # Club all normal NV samples together
    grepl("^NV", sampleNames) ~ "NV",

    # Keep follow up sample separate
    grepl("EC2_FU", sampleNames) ~ 
      str_extract(sampleNames, "EC2_FU"),

    # Keep EC samples separate 
    grepl("^EC\\d+", sampleNames) ~ 
      str_extract(sampleNames, "^EC\\d+"),

    # Keep CAR-T controls separate
    grepl("^CTRL\\d+", sampleNames) ~ 
      str_extract(sampleNames, "^CTRL\\d+"),

    TRUE ~ NA_character_
  ),
  sampleNames
)

# Apply category to Seurat object
combined_data$category <- plyr::mapvalues(
  combined_data$sample_names,
  from = names(sample_to_category),
  to = sample_to_category
)

unique(combined_data$category)
```


```{r chemistry_annotation, message=FALSE, warning=FALSE}
Idents(combined_data) <- "sample_names"

cluster_annotations <- c(
  "EC1"   = "3prime",  "EC2" = "3prime",
  "EC2_FU" = "3prime",  "EC3" = "5prime",
  "EC4"   = "5prime",  "NV1"  = "3prime",
  "NV2" = "3prime",  "NV3" = "3prime",
  "NV4" = "3prime",  "NV5" = "3prime",
  "NV6" = "3prime",  "NV7" = "3prime",
  "NV8" = "3prime",  "NV9" = "3prime",
  "NV10" = "3prime",  "NV11" = "3prime",
  "CTRL2" = "3prime",  "CTRL3" = "3prime",
  "CTRL4" = "5prime",  "CTRL5" = "5prime",
  "CTRL7"  = "5prime"
)

combined_data$chemistry <- plyr::mapvalues(
  combined_data$sample_names,
  from = names(cluster_annotations),
  to   = cluster_annotations
)
unique(combined_data$chemistry)

saveRDS(combined_data, "RDS/data_combined_raw.RDS")
```


---

## Doublet detection and removal (scDblFinder and scds)

```{r scdblfinder, message=FALSE, warning=FALSE}
set.seed(100)

DefaultAssay(combined_data) <- "RNA"

sce <- combined_data %>% JoinLayers() %>% as.SingleCellExperiment()
sce <- scDblFinder(sce, samples = "sample_names")

table(sce$scDblFinder.class)
```

```{r scdblfinder_transfer, message=FALSE, warning=FALSE}
# Transfer scDblFinder scores back to Seurat metadata
meta_scdblfinder <- sce@colData@listData %>%
  as.data.frame() %>%
  dplyr::select(starts_with("scDblFinder"))

rownames(meta_scdblfinder) <- sce@colData@rownames

combined_data <- AddMetaData(
  object   = combined_data,
  metadata = meta_scdblfinder %>% dplyr::select("scDblFinder.class")
)

# QC tracking table
qc_table <- data.frame(
  sample    = unique(combined_data$sample_names),
  cells_raw = as.numeric(table(combined_data$sample_names)[unique(combined_data$sample_names)]),
  stringsAsFactors = FALSE
)

add_qc_column <- function(qc_table, seurat_obj, col_name) {
  counts <- table(seurat_obj$sample_names)
  qc_table[[col_name]] <- counts[qc_table$sample]
  qc_table[[col_name]][is.na(qc_table[[col_name]])] <- 0
  qc_table
}
```


```{r scds, message=FALSE, warning=FALSE}
set.seed(1234)

# Subset to scDblFinder singlets first
data_singlet <- subset(combined_data, scDblFinder.class == "singlet")
qc_table     <- add_qc_column(qc_table, data_singlet, "after_scDblFinder")

# Run scds
sce_scds <- data_singlet %>% JoinLayers() %>% as.SingleCellExperiment()
sce_scds <- cxds(sce_scds)
sce_scds <- bcds(sce_scds)
sce_scds <- cxds_bcds_hybrid(sce_scds, estNdbl = TRUE)

scds_res <- as.data.frame(colData(sce_scds)) %>%
  dplyr::select(cxds_score, bcds_score, hybrid_score, cxds_call, bcds_call, hybrid_call)

data_singlet <- AddMetaData(data_singlet, scds_res)

data_singlet$doublet_consensus <- data_singlet$scDblFinder.class == "doublet" &
                                   data_singlet$hybrid_call == TRUE

table(data_singlet$doublet_consensus)
```

```{r scds_filter, message=FALSE, warning=FALSE}
data_singlet2 <- subset(data_singlet, hybrid_call == FALSE)
qc_table      <- add_qc_column(qc_table, data_singlet2, "after_scds")

saveRDS(data_singlet2, "RDS/data_dblt_filt_combined_raw.RDS")
```

---

## Selection of per-sample QC thresholds


```{r mito, message=FALSE, warning=FALSE}
data <- readRDS("RDS/data_dblt_filt_combined_raw.RDS")

data[["percent.mt"]] <- PercentageFeatureSet(data, pattern = "^MT-")
VlnPlot(data, features = "percent.mt")

data$mt_label <- factor(ifelse(data$percent.mt > 20, "Above 20%", "Below 20%"))

VlnPlot(data, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3)
```

---

# Normalization, Scaling, Finding HVGs, PCA and UMAP 

```{r ig_exclusion_hvg, message=FALSE, warning=FALSE}
data_copy <- data

# Identify and remove IG genes from HVG candidates
ig_genes  <- grep("^IG[HKL][VDJC]", rownames(data_copy), value = TRUE)
genes_keep <- setdiff(rownames(data_copy), ig_genes)

data_no_ig <- subset(data_copy, features = genes_keep)
data_no_ig <- NormalizeData(data_no_ig)
data_no_ig <- FindVariableFeatures(data_no_ig, selection.method = "vst", nfeatures = 2000)

hvg_no_ig <- VariableFeatures(data_no_ig)
VariableFeatures(data) <- hvg_no_ig

data <- NormalizeData(data)
data <- ScaleData(data, vars.to.regress = "percent.mt", verbose = TRUE)
data <- RunPCA(data, npcs = 50, features = VariableFeatures(data))
ElbowPlot(data, ndims = 50)
data <- RunUMAP(data, reduction = "pca", dims = 1:30)

saveRDS(data, "RDS/data_combined_raw_umap.RDS")
```

---

## Density plots to determine per sample QC

```{r density_plots, message=FALSE, warning=FALSE}
# Density plots were generated at this stage to examine the per-sample nCount_RNA distributions. Based on these distributions, sample-specific minimum nCount_RNA thresholds were set for CTRL3 (nCount_CTRL3 = 3500) and NV5 (nCount_NV5 = 1650) in the params block above. All QC filtering was redone to follow these cutoffs and checked again via density pltos

get_data <- function(fn) {
  dat <- get(fn)
  data.table(
    nFeature  = dat$nFeature_RNA,
    percent.mt = dat$percent.mt,
    nCount    = dat$nCount_RNA,
    sample    = dat$sample_names
  )
}

get_cumperc <- function(x) {
  ret <- data.table(percent = seq(0, 100))
  ret[, cells := 0]
  for (i in 1:nrow(ret)) {
    set(ret, i, "cells", 100 * length(which(x > ret$percent[i])) / length(x))
  }
  ret
}

dat <- get_data("data")

dat_filt <- dat[
  (sample != "CTRL3" & sample != "NV5" & nCount > nCount_min) |
  (sample == "CTRL3" & nCount > nCount_CTRL3) |
  (sample == "NV5"    & nCount > nCount_NV5)
]
dat_filt <- dat_filt[nFeature > nFeature_min & nFeature < nFeature_max &
                     percent.mt < percent_mt_max & nCount < nCount_max]

distinct <- scales::hue_pal()(length(unique(dat$sample)))

density_nofilt <- ggplot(dat, aes(x = nCount, colour = sample)) +
  geom_freqpoly(binwidth = 1000) +
  scale_x_continuous(breaks = seq(0, 1e6, by = 1e4)) +
  scale_colour_manual(values = distinct) +
  scale_y_continuous(breaks = seq(0, 3400, by = 500)) +
  theme_minimal() +
  ggtitle("Density Pre-filter") +
  theme(axis.text.x = element_text(angle = 45))
print(density_nofilt)

density_filt <- ggplot(dat_filt, aes(x = nCount, colour = sample)) +
  geom_freqpoly(binwidth = 1000) +
  scale_x_continuous(breaks = seq(0, 1e6, by = 1e4)) +
  scale_colour_manual(values = distinct) +
  theme_minimal() +
  ggtitle("Density Post-filter") +
  theme(axis.text.x = element_text(angle = 45, hjust = 0.5, vjust = 0.5))
print(density_filt)

all_cum <- dat[, get_cumperc(percent.mt), by = "sample"]; all_cum$filter <- "no"
fil_cum <- dat_filt[, get_cumperc(percent.mt), by = "sample"]; fil_cum$filter <- "yes"

percent.mt_plt <- ggplot(rbind(all_cum, fil_cum),
       aes(x = percent, y = cells, color = sample, lty = filter)) +
  geom_line() +
  xlab("> percent.mt") + ylab("Percent of cells") +
  scale_x_continuous(breaks = seq(0, 100, by = 10)) +
  scale_colour_manual(values = distinct) +
  theme_minimal() +
  ggtitle("Cumulative Mitochondrial Percentage") +
  theme(axis.text.x = element_text(angle = 45))
print(percent.mt_plt)
```

---

## Cell Filtering

```{r cell_filtering, message=FALSE, warning=FALSE}
data <- readRDS("RDS/data_combined_raw_umap.RDS")

current <- data

current <- subset(current, subset = nFeature_RNA > nFeature_min)
qc_table$nFeature_RNA_200 <- as.numeric(table(current$sample_names)[qc_table$sample])

current <- subset(current, subset = nFeature_RNA < nFeature_max)
qc_table$nFeature_RNA_8000 <- as.numeric(table(current$sample_names)[qc_table$sample])

current <- subset(current, subset = percent.mt < percent_mt_max)
qc_table$percent.mt_20 <- as.numeric(table(current$sample_names)[qc_table$sample])

current <- subset(
  current,
  subset =
    (sample_names != "CTRL3" & sample_names != "NV5" & nCount_RNA > nCount_min) |
    (sample_names == "CTRL3" & nCount_RNA > nCount_CTRL3) |
    (sample_names == "NV5"    & nCount_RNA > nCount_NV5)
)
qc_table$nCount_1000 <- as.numeric(table(current$sample_names)[qc_table$sample])

current <- subset(current, subset = nCount_RNA < nCount_max)
qc_table$nCount_100k <- as.numeric(table(current$sample_names)[qc_table$sample])

# Apply final filter to data object
data_filtered <- subset(
  data,
  subset =
    nFeature_RNA > nFeature_min &
    nFeature_RNA < nFeature_max &
    percent.mt   < percent_mt_max &
    nCount_RNA   < nCount_max &
    (
      (sample_names != "CTRL3" & sample_names != "NV5" & nCount_RNA > nCount_min) |
      (sample_names == "CTRL3" & nCount_RNA > nCount_CTRL3) |
      (sample_names == "NV5"    & nCount_RNA > nCount_NV5)
    )
)

saveRDS(data_filtered, "RDS/data_filtered_umap.RDS")
```

---

## Non-integrated UMAP Before Filtering Post Doublet Filtering {.tabset}

```{r umap_before_qc, results='asis', message=FALSE, warning=FALSE}
data <- readRDS("RDS/data_combined_raw_umap.RDS")

cat("## Samples \n")
DimPlot(data, reduction = "umap", group.by = "sample_names", raster = FALSE) +
  ggtitle("Before Integration and QC") + theme(legend.position = "none")
cat(" \n \n")

cat("## Condition \n")
DimPlot(data, reduction = "umap", group.by = "condition", raster = FALSE) +
  ggtitle("Before Integration and QC")
cat(" \n \n")
```


---

## Non-integrated UMAP After Filtering {.tabset}

```{r umap_after_filter, results='asis', message=FALSE, warning=FALSE}
cat("## Samples \n")
DimPlot(data_filtered, reduction = "umap", group.by = "sample_names", raster = FALSE) +
  ggtitle("UMAP post Mitochondrial Filtering") + theme(legend.position = "none")
cat(" \n \n")

cat("## Condition \n")
DimPlot(data_filtered, reduction = "umap", group.by = "condition", raster = FALSE) +
  ggtitle("UMAP post Mitochondrial Filtering")
cat(" \n \n")
```

---

## QC Table

```{r qc_table_save, message=FALSE, warning=FALSE}
qc_table[nrow(qc_table) + 1, ] <- NA
qc_table$sample[nrow(qc_table)] <- "Total"

num_cols <- sapply(qc_table, is.numeric)
qc_table[nrow(qc_table), num_cols] <- colSums(qc_table[, num_cols], na.rm = TRUE)

openxlsx::write.xlsx(
  qc_table,
  "csvs/all_celltypes_qc_table_cell_counts.xlsx",
  rowNames = FALSE
)
```

```{r qc_table_display, results='asis', message=FALSE, warning=FALSE}
print(qc_table)
```

